---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r, echo=FALSE, include=FALSE}
setwd("C:/Users/rossd/OneDrive/Documents/GitHub/RepData_PeerAssessment1")
unzip("activity.zip")
data <- read.csv("activity.csv")
require(tidyverse)
```



## What is mean total number of steps taken per day?
```{r, echo=FALSE}

daily.steps<-group_by(data, date)%>%
  summarize(Steps = sum(steps, na.rm = TRUE))
data.na.rm <- filter(data, !is.na(steps))
mean.steps <- mean(daily.steps$Steps)
median.steps <- median(daily.steps$Steps)
hist.data <- data.frame(date = as.Date(rep(data.na.rm$date, data.na.rm$steps)))  
ggplot(hist.data, aes(x = date))+
  geom_histogram(binwidth = 1)+
  ggtitle("Steps taken per day") +
  xlab("Date") +
  ylab("Number of Steps")
```
The figure above displays the absolute number of steps recorded each day. The mean number of steps per day is `r mean.steps` while the median number is `r median.steps`


## What is the average daily activity pattern?
```{r, echo=FALSE}
interval.data <- group_by(data, interval)%>%
  summarize(steps = mean(steps, na.rm = TRUE))
max.interval.steps <- interval.data[which(interval.data$steps == max(interval.data$steps)),]
plot(x = interval.data$interval, y = interval.data$steps,
     type = "l",
     main = "Steps Over Time averaged across Days",
     sub = "Recorded in 5 min intervals",
     xlab = "Minute",
     ylab = "Number of Steps")
```
  The figure above displays the average number of steps recorded at different times of the day. The largest average number of steps per five minute interval is `r max.interval.steps$steps[[1]]` recorded at `r max.interval.steps$interval[[1]]`


## Imputing missing values
```{r, echo=FALSE}
missing.value.sum <- sum(is.na(data$steps))
interval.average.data <- group_by(data, interval)%>%
  summarize(steps = median(steps, na.rm = TRUE))
missing.values <- which(is.na(data$steps))
imputed.data <- data
for (i in missing.values) {
  imputed.data$steps[i]<- interval.average.data$steps[interval.average.data$interval==imputed.data$interval[i]]
  
}

imputed.hist.data <- data.frame(date = as.Date(rep(imputed.data$date, imputed.data$steps)))  
ggplot(imputed.hist.data, aes(x = date))+
  geom_histogram(binwidth = 1)+
  ggtitle("Steps taken per day", subtitle = "missing data imputed from median at that time interval") +
  xlab("Date") +
  ylab("Number of Steps")

imputed.daily.steps<-group_by(imputed.data, date)%>%
  summarize(Steps = sum(steps, na.rm = FALSE))

imputed.mean.steps <- mean(imputed.daily.steps$Steps)
imputed.median.steps <- median(imputed.daily.steps$Steps)

```
  The figure above mimics the first figure displayed in this paper with one key difference: missing values have been imputed using the median number of steps recorded at that time of day. This has shifted the mean steps taken per day from `r mean.steps` to `r imputed.mean.steps` and the median from `r median.steps` to `r imputed.median.steps`


## Are there differences in activity patterns between weekdays and weekends?

```{r, echo=FALSE}
weekday.data <- imputed.data
weekday.data$date <- as.Date(weekday.data$date)
weekday.data$weekday <- case_when(weekdays(weekday.data$date) == "Saturday" ~ "Weekend",
                                  weekdays(weekday.data$date) == "Sunday" ~ "Weekend",
                                  TRUE ~ "Weekday") %>%
  as.factor()
weekend.data <- weekday.data %>%
  filter(weekday == "Weekend") %>%
  group_by(interval)%>%
  summarize(steps = mean(steps))
weekday.data <- weekday.data %>%
  filter(weekday == "Weekday")%>%
  group_by(interval)%>%
  summarize(steps = mean(steps))

par(mfrow = c(2,1))
plot(weekday.data$interval, weekday.data$steps,
     main = "Average Steps Taken",
     sub = "Weekdays Only",
     type= "l",
     xlab = "Time Interval",
     ylab = "Steps Taken")
plot(weekend.data$interval, weekend.data$steps,
     main = "Average Steps Taken",
     sub = "Weekends Only",
     type = "l",
     xlab = "Time Interval",
     ylab = "Steps Taken")
```

